Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2015-08-24 Rmarkdown/biology_background.Rmd
Change some tables to figures (closes #14).

Whole-genome bisulfite-sequencing remains an expensive assay, which limits its use in studies involving large numbers of samples. Furthermore, depending on the choice of several sequencing parameters, approximately $35 \%$ to $72\%$ of reads will not contain any CpGs (Table \ref{tab:reads_without_cpgs}), which might be considered a gross waste of resources for some experiments.

\begin{table}[h]
\centering
\caption[Percentage of reads without CpGs]{The percentage of reads that do not contain any CpGs for each sample in the \emph{Lister} dataset. Paired-end reads are counted as a single read and as such have a lower percentage of reads without CpGs. See Chapter \ref{chap:datasets} for a description of the \emph{Lister} dataset.}
\label{tab:reads_without_cpgs}
\begin{tabular}{@{}rc@{}}
\toprule
Sample              & Reads with no CpG \\ \midrule
ADS-iPSC            & $35 \%$                \\
ADS-adipose         & $37 \%$                \\
ADS                 & $39 \%$                \\
H9\_Laurent         & $39 \%$                \\
FF-iPSC\_19.11      & $48 \%$                \\
FF-iPSC\_19.7       & $49 \%$                \\
H1+BMP4             & $49 \%$                \\
FF-iPSC\_6.9        & $50 \%$                \\
IMR90-iPSC          & $50 \%$                \\
H9                  & $51 \%$                \\
FF-iPSC\_19.11+BMP4 & $52 \%$                \\
FF                  & $53 \%$                \\
IMR90\_r2           & $57 \%$                \\
IMR90\_r1           & $58 \%$                \\
H1\_r2              & $59 \%$                \\
H1\_r1              & $62 \%$                \\
HSF1                & $72 \%$                \\ \bottomrule
\end{tabular}
\end{table}

@@

Another disadvantage of bisulfite-conversion assays is that they require knowledge of the underlying DNA sequence in order to infer the methylation states of cytosines. This requires either a parallel experiment to sequence the target region(s) or reliance on a reference genome. When relying on a reference genome, the inference of the methylation state can be confounded by the DNA sequence of the sample \citep{Liu:2012ge}. Figure \ref{fig:bisulfite_CT_SNP.pdf} illustrates such an example.

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/bisulfite_CT_SNP.pdf}
\caption[Sequence variation confounds methylation calling]{Sequence variation between the reference genome and the sample's genome can result in incorrect inference about the methylation state of the sample's genome. The locus in orange is a cytosine in the reference genome but a thymine in the sample's genome. Because the read is compared against the reference genome, it may incorrectly inferred to be an unmethylated cytosine. The locus in purple is a thymine in the reference genome but an unmethylated cytosine in the sample's genome. Because the read is compared against the reference genome, it may incorrectly inferred to be a thymine.}
\label{fig:bisulfite_CT_SNP.pdf}
\end{figure}

>>>>

Whole-genome bisulfite-sequencing remains an expensive assay, which limits its use in studies involving large numbers of samples. Furthermore, depending on the choice of several sequencing parameters, approximately $35 \%$ to $72\%$ of reads will not contain any CpGs (Figure \ref{fig:reads_without_cpgs}), which might be considered a gross waste of resources for some experiments.

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/percentage-reads-without-CpGs.pdf}
\caption[Percentage of reads without CpGs]{The percentage of reads that do not contain any CpGs for each sample in the \emph{Lister} dataset. Paired-end (PE) reads are counted as a single unit. Therefore, samples sequenced with paired-end reads have a lower percentage of reads without CpGs than do samples sequenced with single-end (SE) reads. See Chapter \ref{chap:datasets} for a description of the \emph{Lister} dataset.}
\label{fig:reads_without_cpgs}
\end{figure}

@@

Another disadvantage of bisulfite-conversion assays is that they require knowledge of the underlying DNA sequence in order to infer the methylation states of cytosines. This requires either a parallel experiment to sequence the target region(s) or reliance on a reference genome. When relying on a reference genome, the inference of the methylation state can be confounded by the DNA sequence of the sample \citep{Liu:2012ge}. Figure \ref{fig:bisulfite_CT_SNP} illustrates such an example.

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/bisulfite_CT_SNP.pdf}
\caption[Sequence variation confounds methylation calling]{Sequence variation between the reference genome and the sample's genome can result in incorrect inference about the methylation state of the sample's genome. The locus in orange is a cytosine in the reference genome but a thymine in the sample's genome. Because the read is compared against the reference genome, it may incorrectly inferred to be an unmethylated cytosine. The locus in purple is a thymine in the reference genome but an unmethylated cytosine in the sample's genome. Because the read is compared against the reference genome, it may incorrectly inferred to be a thymine.}
\label{fig:bisulfite_CT_SNP}
\end{figure}

### 2015-08-24 Rmarkdown/datasets.Rmd
Change some tables to figures (closes #14).

\begin{table}[!htb]
\centering
\caption[Description of \emph{Lister} dataset part 1]{Samples in \emph{Lister} dataset from \cite{Lister:2009hy}. The reported read lengths are prior to any trimming of the reads. Abbreviations: \emph{ESC} = embryonic stem cell; \emph{SE} = single-end sequencing.}
\label{tab:Lister_dataset_pt1}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Sample              & Tissue type             & Sequencing & Read length & Ave. coverage \\
IMR90\_r1           & Lung fibroblasts        & SE         & $87$ bp    & $14\times$ \\
IMR90\_r2           & Lung fibroblasts        & SE         & $87$ bp    & $15\times$ \\
H1\_r1              & ESC                     & SE         & $85$ bp    & $15\times$ \\
H1\_r2              & ESC                     & SE         & $85$ bp    & $14\times$ \\
\bottomrule
\end{tabulary}
\end{table}

@@

\begin{table}[!htb]
\centering
\caption[Description of \emph{Lister} dataset part 2]{Samples in \emph{Lister} dataset from \cite{Lister:2011kg}. The reported read lengths are prior to any trimming of the reads. Abbreviations: \emph{iPSC} = induced pluripotent stem cell; \emph{ESC} = embryonic stem cell; \emph{IVD} = \emph{in vitro} differentiated from pluripotent cell line; \emph{SE} = single-end sequencing; \emph{PE} = paired-end sequencing.}
\label{tab:Lister_dataset_pt2}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Sample              & Tissue type             & Sequencing & Read length & Ave. coverage \\
ADS                 & Adipose                 & PE         & $75$ bp    & $23\times$ \\
ADS-adipose         & Adipocytes from ADS     & PE         & $75$ bp    & $24\times$ \\
ADS-iPSC            & iPSC from ADS           & PE         & $75$ bp    & $26\times$ \\
FF                  & Foreskin fibroblasts    & SE         & $85$ bp    & $16\times$ \\
FF-iPSC\_6.9        & iPSC from FF            & SE         & $85$ bp    & $10\times$ \\
FF-iPSC\_19.7       & iPSC from FF            & SE         & $85$ bp    & $9\times$  \\
FF-iPSC\_19.11      & iPSC from FF            & SE         & $85$ bp    & $8\times$  \\
FF-iPSC\_19.11+BMP4 & IVD from FF-iPSC\_19.11 & SE         & $85$ bp    & $17\times$ \\
IMR90-iPSC          & iPSC from IMR90         & SE         & $85$ bp    & $9\times$ \\
H1+BMP4             & IVD from H1             & SE         & $85$ bp    & $33\times$ \\
H9                  & ESC                     & SE         & $85$ bp    & $9\times$ \\
H9\_Laurent         & ESC                     & PE         & $75$ bp    & $8\times$ \\
HSF1                & ESC                     & SE         & $47$ bp    & $5\times$ \\
\bottomrule
\end{tabulary}
\end{table}

@@

\begin{table}[!htb]
\centering
\caption[Description of \emph{EPISCOPE} dataset]{Samples in \emph{EPISCOPE} dataset. The reported read lengths are prior to any trimming of the reads.}
\label{tab:EPISCOPE_dataset}
\begin{tabulary}{\textwidth}{LLLL}
\toprule
Sample & Sequencing & Read length & Ave. coverage \\ \midrule
E13BUF & PE         & $101$ bp    & $8\times$     \\
E13SA  & PE         & $101$ bp    & $28\times$    \\
E13VA  & PE         & $150$ bp    & $27\times$    \\
E13VAT & PE         & $101$ bp    & $25\times$    \\
E18BUF & PE         & $101$ bp    & $21\times$    \\
E18SA  & PE         & $101$ bp    & $25\times$    \\
E18VA  & PE         & $150$ bp    & $36\times$    \\
E18VAT & PE         & $101$ bp    & $26\times$    \\
E23BUF & PE         & $101$ bp    & $12\times$    \\
E23SA  & PE         & $101$ bp    & $29\times$    \\
E23VA  & PE         & $101$ bp    & $32\times$    \\
E23VAT & PE         & $101$ bp    & $31\times$    \\ \bottomrule
\end{tabulary}
\end{table}

@@

\begin{table}[!htb]
\centering
\caption[Description of \emph{Seisenberger} dataset]{Samples in \emph{Seisenberger} dataset. The reported read lengths are prior to any trimming of the reads. All samples were first published in \citet{Seisenberger:2012ko}.}
\label{tab:Seisenberger_dataset}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Sample            & Tissue                     & Sequencing & Read length & Ave. coverage             \\ \midrule
J1\_1             & ESC                        & PE         & $100$ bp    & $12\times$     \\
E6.5\_epiblast\_1 & Epiblast                   & PE         & $100$ bp    & $13\times$     \\
E16.5\_male\_1    & Male progenitor germ cells & PE         & $100$ bp    & $12\times$     \\ \bottomrule
\end{tabulary}
\end{table}

@@

\begin{table}[!htb]
\centering
\caption[Description of \emph{Ziller} dataset]{Sequencing runs in \emph{Ziller} dataset. The reported read lengths are prior to any trimming of the reads. All samples were first published in \citet{Ziller:2013fs}.}
\label{tab:Ziller_dataset_sequencing_runs}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Sample    & Tissue                     & Sequencing & Read length & Ave. coverage        \\ \midrule
SRR949193 & Frontal\_cortex\_normal\_1 & PE         & 101 bp         & $10\times$ \\
SRR949194 & Frontal\_cortex\_normal\_1 & PE         & 101 bp         & $10\times$    \\
SRR949195 & Frontal\_cortex\_normal\_1 & PE         & 101 bp         & $10\times$    \\
SRR949196 & Frontal\_cortex\_normal\_2 & PE         & 101 bp         & $9\times$     \\
SRR949197 & Frontal\_cortex\_normal\_2 & PE         & 101 bp         & $9\times$     \\
SRR949198 & Frontal\_cortex\_normal\_2 & PE         & 101 bp         & $9\times$     \\
SRR949199 & Frontal\_cortex\_AD\_1     & PE         & 101 bp         & $9\times$     \\
SRR949201 & Frontal\_cortex\_AD\_1     & PE         & 101 bp         & $9\times$     \\
SRR949202 & Frontal\_cortex\_AD\_2     & PE         & 101 bp         & $10\times$    \\
SRR949203 & Frontal\_cortex\_AD\_2     & PE         & 101 bp         & $10\times$    \\
SRR949206 & HepG2\_cell\_line          & PE         & 101 bp         & $2\times$     \\
SRR949207 & HepG2\_cell\_line          & PE         & 101 bp         & $1\times$     \\
SRR949208 & IMR90\_cell\_line          & PE         & 101 bp         & $1\times$     \\
SRR949209 & IMR90\_cell\_line          & PE         & 101 bp         & $3\times$     \\
SRR949210 & Colon\_Tumor\_Primary      & PE         & 101 bp         & $8\times$     \\
SRR949211 & Colon\_Tumor\_Primary      & PE         & 101 bp         & $8\times$     \\
SRR949212 & Colon\_Tumor\_Primary      & PE         & 101 bp         & $9\times$     \\
SRR949213 & Colon\_Tumor\_Primary      & PE         & 101 bp        & $8\times$     \\
SRR949215 & Colon\_Primary\_Normal     & PE         & 101 bp         & $8\times$     \\ \bottomrule
\end{tabulary}
\end{table}

>>>>

\begin{table}[!htb]
\centering
\caption[Description of \emph{Lister} dataset part 1]{Samples in \emph{Lister} dataset from \cite{Lister:2009hy}. The reported read lengths are prior to any trimming of the reads. Abbreviations: \emph{ESC} = embryonic stem cell; \emph{SE} = single-end sequencing.}
\label{tab:Lister_dataset_pt1}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Sample              & Tissue type             & Sequencing & Read length & Ave. coverage \\ \midrule
IMR90\_r1           & Lung fibroblasts        & SE         & $87$ bp    & $14\times$ \\
IMR90\_r2           & Lung fibroblasts        & SE         & $87$ bp    & $15\times$ \\
H1\_r1              & ESC                     & SE         & $85$ bp    & $15\times$ \\
H1\_r2              & ESC                     & SE         & $85$ bp    & $14\times$ \\
\bottomrule
\end{tabulary}
\end{table}

@@

\begin{table}[!htb]
\centering
\caption[Description of \emph{Lister} dataset part 2]{Samples in \emph{Lister} dataset from \cite{Lister:2011kg}. The reported read lengths are prior to any trimming of the reads. Abbreviations: \emph{iPSC} = induced pluripotent stem cell; \emph{ESC} = embryonic stem cell; \emph{IVD} = \emph{in vitro} differentiated from pluripotent cell line; \emph{SE} = single-end sequencing; \emph{PE} = paired-end sequencing.}
\label{tab:Lister_dataset_pt2}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Sample              & Tissue type             & Sequencing & Read length & Ave. coverage \\ \midrule
ADS                 & Adipose                 & PE         & $75$ bp    & $23\times$ \\
ADS-adipose         & Adipocytes from ADS     & PE         & $75$ bp    & $24\times$ \\
ADS-iPSC            & iPSC from ADS           & PE         & $75$ bp    & $26\times$ \\
FF                  & Foreskin fibroblasts    & SE         & $85$ bp    & $16\times$ \\
FF-iPSC\_6.9        & iPSC from FF            & SE         & $85$ bp    & $10\times$ \\
FF-iPSC\_19.7       & iPSC from FF            & SE         & $85$ bp    & $9\times$  \\
FF-iPSC\_19.11      & iPSC from FF            & SE         & $85$ bp    & $8\times$  \\
FF-iPSC\_19.11+BMP4 & IVD from FF-iPSC\_19.11 & SE         & $85$ bp    & $17\times$ \\
IMR90-iPSC          & iPSC from IMR90         & SE         & $85$ bp    & $9\times$ \\
H1+BMP4             & IVD from H1             & SE         & $85$ bp    & $33\times$ \\
H9                  & ESC                     & SE         & $85$ bp    & $9\times$ \\
H9\_Laurent         & ESC                     & PE         & $75$ bp    & $8\times$ \\
HSF1                & ESC                     & SE         & $47$ bp    & $5\times$ \\
\bottomrule
\end{tabulary}
\end{table}

@@

\begin{table}[!htb]
\centering
\caption[Description of \emph{EPISCOPE} dataset]{Samples in \emph{EPISCOPE} dataset. The reported read lengths are prior to any trimming of the reads. Abbreviations: \emph{PE} = paired-end sequencing; \emph{BUF} = buffy coat; \emph{SA} = subcutaneous adipose; \emph{VA} = visceral adipocytes; \emph{VAT} = visceral adipose tissue.}
\label{tab:EPISCOPE_dataset}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Sample & Tissue & Sequencing & Read length & Ave. coverage \\ \midrule
E13BUF & BUF & PE         & $101$ bp    & $8\times$     \\
E13SA  & SA  & PE         & $101$ bp    & $28\times$    \\
E13VA  & VA  & PE         & $150$ bp    & $27\times$    \\
E13VAT & VAT & PE         & $101$ bp    & $25\times$    \\
E18BUF & BUF & PE         & $101$ bp    & $21\times$    \\
E18SA  & SA  & PE         & $101$ bp    & $25\times$    \\
E18VA  & VA  & PE         & $150$ bp    & $36\times$    \\
E18VAT & VAT & PE         & $101$ bp    & $26\times$    \\
E23BUF & BUF & PE         & $101$ bp    & $12\times$    \\
E23SA  & SA  & PE         & $101$ bp    & $29\times$    \\
E23VA  & VA  & PE         & $101$ bp    & $32\times$    \\
E23VAT & VAT & PE         & $101$ bp    & $31\times$    \\ \bottomrule
\end{tabulary}
\end{table}

@@

\begin{table}[!htb]
\centering
\caption[Description of \emph{Seisenberger} dataset]{Samples in \emph{Seisenberger} dataset. The reported read lengths are prior to any trimming of the reads. All samples were first published in \citet{Seisenberger:2012ko}. Abbreviations: \emph{PE} = paired-end sequencing}
\label{tab:Seisenberger_dataset}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Sample            & Tissue                     & Sequencing & Read length & Ave. coverage             \\ \midrule
J1\_1             & ESC                        & PE         & $100$ bp    & $12\times$     \\
E6.5\_epiblast\_1 & Epiblast                   & PE         & $100$ bp    & $13\times$     \\
E16.5\_male\_1    & Male progenitor germ cells & PE         & $100$ bp    & $12\times$     \\ \bottomrule
\end{tabulary}
\end{table}

@@

\begin{table}[!htb]
\centering
\caption[Description of \emph{Ziller} dataset]{Sequencing runs in \emph{Ziller} dataset. The reported read lengths are prior to any trimming of the reads. All samples were first published in \citet{Ziller:2013fs}. Abbreviations: \emph{PE} = paired-end sequencing}
\label{tab:Ziller_dataset_sequencing_runs}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Sample    & Tissue                     & Sequencing & Read length & Ave. coverage        \\ \midrule
SRR949193 & Frontal\_cortex\_normal\_1 & PE         & 101 bp         & $10\times$ \\
SRR949194 & Frontal\_cortex\_normal\_1 & PE         & 101 bp         & $10\times$    \\
SRR949195 & Frontal\_cortex\_normal\_1 & PE         & 101 bp         & $10\times$    \\
SRR949196 & Frontal\_cortex\_normal\_2 & PE         & 101 bp         & $9\times$     \\
SRR949197 & Frontal\_cortex\_normal\_2 & PE         & 101 bp         & $9\times$     \\
SRR949198 & Frontal\_cortex\_normal\_2 & PE         & 101 bp         & $9\times$     \\
SRR949199 & Frontal\_cortex\_AD\_1     & PE         & 101 bp         & $9\times$     \\
SRR949201 & Frontal\_cortex\_AD\_1     & PE         & 101 bp         & $9\times$     \\
SRR949202 & Frontal\_cortex\_AD\_2     & PE         & 101 bp         & $10\times$    \\
SRR949203 & Frontal\_cortex\_AD\_2     & PE         & 101 bp         & $10\times$    \\
SRR949206 & HepG2\_cell\_line          & PE         & 101 bp         & $2\times$     \\
SRR949207 & HepG2\_cell\_line          & PE         & 101 bp         & $1\times$     \\
SRR949208 & IMR90\_cell\_line          & PE         & 101 bp         & $1\times$     \\
SRR949209 & IMR90\_cell\_line          & PE         & 101 bp         & $3\times$     \\
SRR949210 & Colon\_Tumor\_Primary      & PE         & 101 bp         & $8\times$     \\
SRR949211 & Colon\_Tumor\_Primary      & PE         & 101 bp         & $8\times$     \\
SRR949212 & Colon\_Tumor\_Primary      & PE         & 101 bp         & $9\times$     \\
SRR949213 & Colon\_Tumor\_Primary      & PE         & 101 bp        & $8\times$     \\
SRR949215 & Colon\_Primary\_Normal     & PE         & 101 bp         & $8\times$     \\ \bottomrule
\end{tabulary}
\end{table}

### 2015-08-24 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Change some tables to figures (closes #14).

The maximum memory usage is not strictly proportional to chromosome length. It is instead driven by the number and density of CpGs on the chromosome.  For example, chromosome 19, which has the highest CpG density of all the autosomes in the human genome, requires far more memory than chromosome 18, which has less than half the CpG density of chromosome 19 (see Table \ref{tab:cpg_density_hg19}). The relationship between the maximum memory usage and the tuple size (`-m`) is complex; more data have to be retained as `-m` increases, thus increasing the memory usage, but fewer reads contain tuples of that size and so there aren't as many m-tuples or observations on these to count and retain. Memory usage is therefore relatively constant across values of `-m` for a given chromosome. The obvious exception is for the results labelled `2ac`, which used the `--all-combinations` flag in conjunction with `-m 2`. This means that all 2-tuples with $NIL \geq 0$ were extracted and there are many, many more CpG 2-tuples with $NIL \geq 0$ than there are with $NIL = 0$, hence the increase in memory usage.

@@

@@

```{r autosomal_cpg_density_hg19, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
library(xtable)
x <- read.table("../data/wgbs_bioinformatics_analysis/CpG_density_hg19.txt",
                header = TRUE, stringsAsFactors = FALSE)
xt <- xtable(x, caption = c("Percentage of dinucleotides that are CpGs for each chromosome in the human reference genome (hg19).", "CpG percentage per chromosome in hg19"), label = "tab:cpg_density_hg19")
align(xt) <- rep("c", ncol(x) + 1)
digits(xt) <- 2
print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, include.rownames = FALSE, caption.placement = "top")
```

>>>>

@@

The maximum memory usage is not strictly proportional to chromosome length. It is instead driven by the number and density of CpGs on the chromosome.  For example, chromosome 19, which has the highest CpG density of all the autosomes in the human genome, requires far more memory than chromosome 18, which has less than half the CpG density of chromosome 19 (see Figure \ref{fig:cpg_density_hg19}). The relationship between the maximum memory usage and the tuple size (`-m`) is complex; more data have to be retained as `-m` increases, thus increasing the memory usage, but fewer reads contain tuples of that size and so there aren't as many m-tuples or observations on these to count and retain. Memory usage is therefore relatively constant across values of `-m` for a given chromosome. The obvious exception is for the results labelled `2ac`, which used the `--all-combinations` flag in conjunction with `-m 2`. This means that all 2-tuples with $NIL \geq 0$ were extracted and there are many, many more CpG 2-tuples with $NIL \geq 0$ than there are with $NIL = 0$, hence the increase in memory usage.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/CpG_density_hg19.pdf}
\caption[CpG percentage per chromosome in hg19]{Percentage of dinucleotides that are CpGs for each chromosome in the human reference genome (hg19).}
\label{fig:cpg_density_hg19}
\end{figure}

@@
